Proof nets for first-order additive linear logic

### 2018-08-07 ALL1.tex
minor typo stuff

At the heart of a theory of proof nets is the question of \emph{proof identity}: when are two proofs equivalent? The answer determines which proofs should map onto the same proof net. The introduction of quantifiers creates an interesting issue: if two proofs differ by an immaterial choice of existential witness, should they be equivalent? For example, to prove the sequent $\seq{\exists x.P(x)}{\exists y.\dual P(y)}$ both quantifiers must receive \emph{the same} witness, as in the following two proofs, but \emph{any} witness will do.
\[
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(s)}{\dual P(s)}}{}}}
	\quad
	\stackrel?\equiv
	\quad
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(t)}{\dual P(t)}}{}}}
\]
The issue is more pronounced where quantifiers are \emph{vacuous}, $\exists x.A$ with $x$ not free in $A$. The proofs below left can only be distinguished even syntactically because the $\exists R$-rule makes the instantiating witness explicit. Below right is an interesting intermediate variant: the witness $s$ or $t$ can be observed without explicit annotation in the $\exists R$-rule, but the choice is equally immaterial to the content of the proof as when the quantifier were vacuous.
\[
	\vc{\infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{\infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}\qquad
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(s)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(t)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
\]
In this paper we will not attempt to settle the question of proof identity. Rather, our two notions of proof net each represent a natural and coherent perspective, at either end of the spectrum. \emph{Witness nets} make all existential witnesses explicit, including those to vacuous quantifiers, rejecting all three equivalences above. \emph{Unification nets} leave all witnesses implicit, thus identifying all proofs modulo witness assignment, and validating all three equivalences.

@@

We will distinguish \defn{strict coalescence} ($\scoal$) for witness nets and \defn{unifying coalescence} ($\ucoal$) for unification nets. We first give an example of the former (writing $Pxy$ for $P(x,y)$ to conserve space). In the initial witness net, below left, each link corresponds to a sequent calculus axiom between both linked subformulas, with the substitutions applied.
%
\newcommand\xs{xs} %{(\kern-1ptx\kern-1pt,\kern-2pts\kern-1pt)}
\newcommand\xt{xt} %{(\kern-1ptx\kern-1pt,\kern-2ptt\kern-1pt)}
\newcommand\xy{xy} %(\kern-1ptx\kern-1pt,\kern-2pty\kern-1pt)}
\newcommand\zs{zs}
\newcommand\zt{zt}
\[
\hspace*{-\parindent}
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z,s/y]~~$}]{[-3]3,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,s/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \quad\scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,s/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \scoal &
\hfill
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z]$}]{[2]2,4}
	\end{tikzpicture}}
\hfill
\scoal\cdots
\\ \\
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual P\xs}{P\xs}}{}
       &&
	  \color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}
	 }}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{white}\infer{\vphantom{\dual P}}{\color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}}}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{blue}\infer[\!\Qrr?t]{\seq{\exists y.\dual P\xy}{P\xt}}{\color{black}\infer{\seq{\dual P\xt}{P\xt}}{}}}
\end{array}
\]
The first two steps (above middle and right) move both links from the subformula $\dual Pxy$ to $\exists y.\dual Pxy$, removing the substitutions $[s/y]$ and $[t/y]$ on $y$, and corresponding to sequent rules $\exists R,s$ and $\exists R,t$. Observe that we maintain the domain of the substitutions on a link as the variables of those existential quantifiers that either linked subformula is (strictly) in scope of.

@@

Since we will be working with a graphical representation, we will adopt \defn{Barendregt's convention}, that bound variable names are globally unique identifiers, in the following form. In a sequent $\seq AB$, we assume all quantifiers to have a unique binding variable, distinct from any free variable. In a proof $\pi$ over $\seq AB$, a variable $x$ that is universally quantified as $\forall x.C$ in $\seq AB$ is an \defn{eigenvariable}. A $\qrr!$ rule on $\forall x.C$ is considered to bind all free occurrences of the eigenvariable $x$ in its direct subproof. Accordingly, we assume that $x$ does not occur free outside these subproofs (which can be guaranteed by globally renaming $x$). We take variable names to be persistent throughout a proof, in the sense that we don't admit alpha-conversion between proof rules. We thus have unique bound variable names in $\seq AB$, but in $\pi$ all rules on the subformula $\forall x.C$ share the same eigenvariable $x$.

>>>>

At the heart of a theory of proof nets is the question of \emph{proof identity}: when are two proofs equivalent? The answer determines which proofs should map onto the same proof net. The introduction of quantifiers creates an interesting issue: if two proofs differ by an immaterial choice of existential witness, should they be equivalent? For example, to prove the sequent $\seq{\exists x.P(x)}{\exists y.\dual P(y)}$ both quantifiers must receive \emph{the same} witness, as in the following two proofs, but \emph{any} witness will do.
\[
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(s)}{\dual P(s)}}{}}}
	\quad
	\stackrel?\equiv
	\quad
	\vc{\infer={\seq{\exists x.P(x)}{\exists y.\dual P(y)}}{\infer{\seq{P(t)}{\dual P(t)}}{}}}
\]
The issue is more pronounced where quantifiers are \emph{vacuous}, $\exists x.A$ with $x$ not free in $A$. The proofs below left can only be distinguished even syntactically because the $\exists\mathrm R$ rule makes the instantiating witness explicit. Below right is an interesting intermediate variant: the witness $s$ or $t$ can be observed without explicit annotation in the $\exists\mathrm R$ rule, but the choice is equally immaterial to the content of the proof as when the quantifier were vacuous.
\[
	\vc{\infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{\infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P}{\dual P}}{\infer{\seq P{\dual P}}{}}}\qquad
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?s}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(s)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
	\quad
	\stackrel?\equiv
	\quad	
	\vc{
	 \infer[\!\!\scriptstyle{\qrr?t}]{\seq{\exists x.P\+Q(x)}{\dual P}}{
	  \infer[\!\!\scriptstyle{\qrr+1}]{\seq{P\+Q(t)}{\dual P}}{\infer{\seq P{\dual P}}{}}
	}}
\]
In this paper we will not attempt to settle the question of proof identity. Rather, our two notions of proof net each represent a natural and coherent perspective, at either end of the spectrum. \emph{Witness nets} make all existential witnesses explicit, including those to vacuous quantifiers, rejecting all three equivalences above. \emph{Unification nets} leave all witnesses implicit, thus identifying all proofs modulo witness assignment, and validating all three equivalences.

@@

We will distinguish \defn{strict coalescence} ($\scoal$) for witness nets and \defn{unifying coalescence} ($\ucoal$) for unification nets. We first give an example of the former (writing $Pxy$ for $P(x,y)$ to conserve space). In the initial witness net, below left, each link corresponds to a sequent calculus axiom between both linked subformulas, with the substitutions applied.
%
\newcommand\xs{xs} %{(\kern-1ptx\kern-1pt,\kern-2pts\kern-1pt)}
\newcommand\xt{xt} %{(\kern-1ptx\kern-1pt,\kern-2ptt\kern-1pt)}
\newcommand\xy{xy} %(\kern-1ptx\kern-1pt,\kern-2pty\kern-1pt)}
\newcommand\zs{zs}
\newcommand\zt{zt}
\[
\hspace*{-\parindent}
\begin{array}{@{}c@{}c@{}c@{}c@{}c@{}}
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z,s/y]~~$}]{[-3]3,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,s/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \quad\scoal &
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z,s/y]$}]{[1]3,4}
	\end{tikzpicture}}
& \scoal &
\hfill
	\vc{\begin{tikzpicture}[net]
		\formula[y=1]{\forall*{x.}\exists*{y.}{\dual P}*\xy}
		\formula[y=0]{\exists*{z.}P*\zs\*P*\zt}
		\Vlink[red,l,label={$\scriptstyle[x/z]~$}]{[-2]2,2}
		\Vlink[blue,r,label={$~~\scriptstyle[x/z]$}]{[2]2,4}
	\end{tikzpicture}}
\hfill
\scoal\cdots
\\ \\
\text{\scriptsize
	\color{white}
	 \infer{\vphantom{\dual P}}{
	  \color{red}\infer{\seq{\dual P\xs}{P\xs}}{}
       &&
	  \color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}
	 }}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{white}\infer{\vphantom{\dual P}}{\color{blue}\infer{\seq{\dual P\xt}{P\xt}}{}}}
& \hspace*{26pt} &
\text{\scriptsize
	\color{red}\infer[\!\Qrr?s]{\seq{\exists y.\dual P\xy}{P\xs}}{\color{black}\infer{\seq{\dual P\xs}{P\xs}}{}}
~~
	\color{blue}\infer[\!\Qrr?t]{\seq{\exists y.\dual P\xy}{P\xt}}{\color{black}\infer{\seq{\dual P\xt}{P\xt}}{}}}
\end{array}
\]
The first two steps (above middle and right) move both links from the subformula $\dual Pxy$ to $\exists y.\dual Pxy$, removing the substitutions $[s/y]$ and $[t/y]$ on $y$, and corresponding to sequent rules $\qrr?s$ and $\qrr?t$. Observe that we maintain the domain of the substitutions on a link as the variables of those existential quantifiers that either linked subformula is (strictly) in scope of.

@@

Since we will be working with a graphical representation, we will adopt \defn{Barendregt's convention}, that bound variable names are globally unique identifiers, in the following form. In a sequent $\seq AB$ we assume all quantifiers to have a unique binding variable, distinct from any free variable. In a proof $\pi$ over $\seq AB$, a variable $x$ that is universally quantified as $\forall x.C$ in $\seq AB$ is an \defn{eigenvariable}. A $\qrr!$ rule on $\forall x.C$ is considered to bind all free occurrences of the eigenvariable $x$ in its direct subproof. Accordingly, we assume that $x$ does not occur free outside these subproofs (which can be guaranteed by globally renaming $x$). We take variable names to be persistent throughout a proof, in the sense that we don't admit alpha-conversion between proof rules. We thus have unique bound variable names in $\seq AB$, but in $\pi$ all $\qrr!$ rules on the subformula $\forall x.C$ share the same eigenvariable $x$.
